---
title: "Module 11: Edward R. Tufte"
description: |
  Tufte and C. Minard in R
preview:
author:
  - name: Skylar Walsh
    url: {}
date: 2025-11-03
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
  
My assignment for module # 11 is to review Dr. Piwek's posting on Tufte and C. Minard in R and see if I can generate one of the visualizations he discussed in his post that include:
  
1. Marginal histogram scatter plot
2. Dot-dash plot in ggplot2
3. Dot-dash plot in lattice

Or any one I like the best.

My additional goal for this assignment is to take a practical visualization and try to represent it by the data or what the visualization is trying to capture.

1. Install the packages

In order to run the visualization of my choice, I installed the 'ggthemes' package.

2. Run those packages in the background

The package successfully runs in the background, as well as the 'ggplot2' package that is already installed.

3. The Code

After reading the information from http://motioninsocial.com/tufte/#introduction
  
I chose to replicate the 'Minimal line plot in ggplot2' using the 'ggplot2' & 'ggthemes' packages in 'R - the most powerful open-source statistical programming language' - to replicate the excellent visualization practices developed by Edward Tufte.

```{r}
# Load packages
library(ggplot2)
library(ggthemes)

# Create vector variables for the data
x <- 1967:1977
y <- c(0.5,1.8,4.6,5.3,5.3,5.7,5.4,5,5.5,6,5)

# Create data frame
d <- data.frame(x, y)

# Plot the chart with 'ggplot()' from ggplot 2 package
ggplot(d, aes(x,y)) + geom_line() + geom_point(size=3) + theme_tufte(base_size = 15) +
  theme(axis.title=element_blank()) + geom_hline(yintercept = c(5,6), lty=2) + 
  scale_y_continuous(breaks=seq(1, 6, 1), label=sprintf("$%s",seq(300,400,20))) + 
  scale_x_continuous(breaks=x,label=x) +
  annotate("text", x = c(1977,1977.2), y = c(1.5,5.5), adj=1,  family="serif",
           label = c("Per capita\nbudget expandures\nin constant dollars", "5%"))

```

I chose to reproduce this particular visualization because Piwek (2017) suggests that "We start by plotting the most basic graph from page 65 of The Visual Display of Quantitative Information - a minimal line plot. This one is important because it illustrates the most elemental principle - that of minimalism with reduced 'data-ink'". Piwek (2017) then quotes Tufte (1983) explaining that "the 'data-ink' (total ink used to print the graphic) ratio should be equal to '1 - proportion of graphic that can be erased without loss of data-information'." I found the removal of base R's plots to be interesting and displays a useful feature of 'ggplot' to modify the theme layer.


References:

Edward Tufte, The Visual Display of Quantitative Information (Cheshire, 1983), p. 65

Piwek, L. (2017). Tufte in R. Motioninsocial.com. Retrieved November 3, 2025, from http://motioninsocial.com/tufte/#introduction


